---
title: "Choosing Between getaca and the Alternatives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Between getaca and the Alternatives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(getaca)

options(getaca.cache = file.path(tempdir(), "getaca-alternatives"))
Sys.setenv(GETACA_OFFLINE = "true", NOT_CRAN = "true")
```

Several established approaches solve neighbouring problems, and for a good
number of packages one of them is the better fit. This article describes what
each is built around and where the boundaries fall, so the choice can be made
on the shape of the problem rather than on which package someone met first.

The question `getaca` is built around: a package needs a file it cannot ship,
and every machine running that package should get the same bytes.

## Bundling the data in the package

The first thing to rule in or out. A dataset in `data/` needs no cache, no
network, no checksum and no policy, and every one of those is a moving part
that can fail.

CRAN's size guidance makes this a question of megabytes rather than gigabytes,
and the release cadence question usually settles it before the size does: data
in `data/` release when the package releases. For a reference table that
changes yearly and a package that releases yearly, coupling them is a feature.

Bundle when the data are small, static relative to your release cycle, and
redistributable. Look further when any of the three fails.

## A companion data package

The classic R answer to data that are too large for the main package but still
manageable: ship them as their own package, list it in `Suggests`, and put it
on CRAN, Bioconductor or r-universe.

| | companion data package | `getaca` |
|---|---|---|
| Size | fits a repository | too large to bundle |
| Release cadence | coupled to code releases | independent of them |
| Shape | naturally R objects | any file, any format |
| Granularity | all of it, always | users take what they need |
| License | redistribution permitted | download permitted, redistribution discouraged |
| Installation | the usual package machinery | first use, or an explicit prefetch |
| Offline | works once installed | works once cached |

The licence row is the one that decides it most often. A companion package
redistributes the data, which needs terms that permit redistribution. Plenty of
scientific datasets permit download and discourage or forbid redistribution, and
for those a companion package is not available whatever its other merits.

Granularity is the second. A package needing one of fifteen backbones, at a
gigabyte each, would install fifteen through a companion package. Under
`getaca` a user takes what they use.

A companion package can itself use `getaca`, which is occasionally the right
structure when the companion owns expensive build logic. It is rarely the
first recommendation, since it moves the complexity rather than removing it.

## pins

[pins](https://cran.r-project.org/package=pins) publishes "data sets, models,
and other R objects, making it easy to share them across projects and with
your colleagues", across boards including local folders, Posit Connect and
AWS S3.

The unit is a pin on a board, and the person at the centre of the design is
the one publishing an artefact for others to read. Boards abstract over where
that artefact lives, which is what makes the same code work against a network
drive during development and Posit Connect in production.

Reach for `pins` when the artefact is yours, the audience is a team or an
organisation, the storage backend matters, and writing is part of the workflow.
It covers a range of backends `getaca` has no interest in growing.

Reach for `getaca` when the artefact is a third party's, the audience is
everyone who installs your package, and the declaration has to travel inside
the package so that `R CMD check` and an offline user both behave.

The two coexist. A package can declare its public reference data through
`getaca` and its team's internal model artefacts through `pins` without either
knowing about the other.

## BiocFileCache

[BiocFileCache](https://bioconductor.org/packages/BiocFileCache) "creates a
persistent on-disk cache of files that the user can add, update, and
retrieve", for resources that are costly or difficult to create and for web
resources used across sessions, backed by an SQLite metadata database.

It is the established answer inside Bioconductor, and a Bioconductor package
already carrying that dependency stack has little reason to add another. The
model is a cache the calling package manages: your code decides what to add,
when a cached copy needs updating, and what identity a resource has.

`getaca` makes those decisions from a declaration instead. Identity is
`package / name / version`, resolution runs through a policy, and the
retention sweeps are the package's rather than the caller's. That is a
narrower contract, and it exists because the same behaviour then holds for
every declaring package rather than being reimplemented per package.

Reach for `BiocFileCache` when you are in the Bioconductor ecosystem, when the
resources are ones your code creates rather than ones a publisher versions, or
when you want the cache under your own control.

## pooch

The nearest equivalent outside R, and the one a reader arriving from Python will
already know. [pooch](https://www.fatiando.org/pooch/), "a friend to fetch your data
files", is where a Python package puts this problem: a registry of file names and
hashes, a cache folder under the OS convention, downloaders and processors around
them.

The overlap is the middle of the problem, fetch and hash and cache. The two ends
differ, and both differences are about where the constraints come from.

At the front, a pooch registry is a `registry.txt` shipped as package data, and
`version` is documented as "the version string for your project", naming the
subfolder the cache uses. Data identity is the declaring project's identity, so a
repaired mirror or a fresh upstream cut reaches users when the code does. A `getaca`
declaration versions the data separately, holds several versions at once, names the
head, and can resolve through a remote registry the author keeps, which is what lets
`2026-09` reach an installed copy between releases. `urls` in pooch sets one URL per
file; a `getaca` record takes a list of mirrors and walks it until one answers.
pooch also accepts a `doi:` URL and asks figshare, Zenodo or Dataverse for the
download location as it fetches. `getaca` asks the same three archives in
`registry_draft()`, when the registry is written, and a record then carries the
DOI as what the bytes are cited as and the locations as themselves.

At the back, the constraints are CRAN's. `R CMD check` runs the tests, examples and
vignettes of every package on machines with no network, and `tools::R_user_dir()` is
permitted on condition its contents are actively managed. The check clamp, the three
access helpers and the retention sweeps exist for those two sentences.

## A downloader and a cache directory

The common alternative in practice: `download.file()` or `curl`, a directory
under `tools::R_user_dir()`, and a `file.exists()` check. It is a few dozen
lines and it works.

Two things cost more work than they look. The first is
what happens when the bytes at the URL differ from the bytes you expected, which
covers a truncated transfer, a proxy serving an HTML error page, and a
publisher recutting a file. Each of those reaches your parser as a confusing
error some distance from its cause.

The second is CRAN's requirement to fail gracefully with an informative
message when a resource is unavailable, across tests, examples and vignettes,
on a machine with no network. That is the part that turns a few dozen lines
into a few hundred, and it has to be written again in every package that
depends on external data.

`getaca` is that layer written once. The declaring package supplies a
registry.

## Targets and the workflow tools

`targets` and `drake` manage a computational pipeline, tracking which steps
need to re-run when inputs change. That is a different question from where a
file comes from and whether it is the right file.

They compose. A `targets` pipeline can have a target whose command is
`getaca("wfo", package = "taxify")`, which gives the pipeline a path, and the
path a provenance record:

```{r, eval = FALSE}
library(targets)

list(
  tar_target(backbone_path, getaca::getaca("wfo", package = "taxify"),
             format = "file"),
  tar_target(backbone, read_backbone(backbone_path)),
  tar_target(matched, match_names(species, backbone))
)
```

`format = "file"` makes `targets` watch the returned path, so the pipeline
invalidates downstream targets when the resolved version changes. That is the
combination worth reaching for: `getaca` decides which bytes, `targets`
decides what to recompute.

## renv

`renv` records which package versions an analysis used. A `getaca` pin records
which data versions those packages resolved to. Neither subsumes the other: a
package version does not determine a data version once the package is on the
`current` policy, and a data version says nothing about the code that read it.

```{r, eval = FALSE}
renv::snapshot()
getaca_pin(c("taxify", "otherpkg"))
```

Commit both files. Restoring is `renv::restore()` followed by
`getaca_prefetch()` on a connected machine, after which the analysis runs
offline. See `vignette("policies")` for what a pin holds.

## Decision summary

| If | then |
|---|---|
| the data are small, static and redistributable | put them in `data/` |
| the data are moderate, redistributable and release with your code | a companion data package |
| you are publishing your own artefacts to a team | `pins` |
| you are inside Bioconductor, or caching things your code creates | `BiocFileCache` |
| you are managing which steps re-run | `targets`, with `getaca` supplying the path |
| your package declares a third party's versioned file it cannot ship | `getaca` |

## What getaca deliberately does not do

Naming the boundary is part of choosing. None of these is planned:

- **Credential storage.** A declaration names the environment variable a host
  requires, and getaca reads it at the moment of the request. It holds no
  credentials, reads no `.netrc` and talks to no keyring. See
  `vignette("declaring")`.
- **Cloud storage abstractions.** No S3, no Azure, no board concept. A URL is
  a URL.
- **DOI resolution at fetch time.** A `doi` on a record is what the bytes are
  cited as, and it routes nothing. Resolving one through a repository API on
  every fetch would put a second host, which is not itself mirrorable, in front
  of every retrieval, and would break the mirror loop's assumption that the
  locations it walks are independent sources. The archives are read when the
  registry is written instead: `registry_draft()` covers Zenodo, figshare and
  Dataverse, and emits plain locations.
- **Reading data.** `getaca` returns a path and knows nothing about file
  formats. A processor can unpack an archive; nothing reads its contents.
- **Publishing.** Uploading the data, minting the version, computing the
  checksum in the first place: all outside.
- **Semantic version solving.** Version strings are labels, and the registry
  names the head rather than ranking them.
- **Workflow orchestration.** Composing with `targets` is the answer, rather
  than growing a dependency graph.

The list is short on purpose. Each item is a place where a package could grow
into something else, and the dependency footprint that makes `getaca`
inexpensive to depend on is the thing that would go first.

```{r}
packageDescription("getaca")$Imports
```

## Where to go next

- `vignette("quickstart")` for the whole cycle in one pass

- `vignette("declaring")` for what a declaring package writes

- `vignette("policies")` for reproducibility across machines and time
