---
title: "The Cache"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Cache}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(getaca)

# The session temp directory carries the account name of whoever knits, so
# printing a cache path would put it in the built article. Redact it.
local({
  roots <- unique(c(tempdir(),
                    normalizePath(tempdir(), winslash = "\\", mustWork = FALSE),
                    normalizePath(tempdir(), winslash = "/", mustWork = FALSE)))
  roots <- unique(c(gsub("\\", "\\\\", roots, fixed = TRUE), roots))
  roots <- roots[order(nchar(roots), decreasing = TRUE)]
  render <- knitr::knit_hooks$get("output")
  knitr::knit_hooks$set(output = function(x, options) {
    for (root in roots) x <- gsub(root, "<tempdir>", x, fixed = TRUE)
    render(x, options)
  })
})

options(getaca.cache = file.path(tempdir(), "getaca-cache-vignette"))
Sys.setenv(GETACA_OFFLINE = "true", NOT_CRAN = "true")

reg <- registry("yourpkg", list(
  resource("backbone", "2026-06",
           urls = "https://host.invalid/backbone-2026-06.zip",
           sha256 = strrep("9f", 32), size = 797e6, license = "CC-BY-4.0")
))
```

The cache is where a declared resource becomes a local path. This article
covers the layout, what makes a cached copy trustworthy, how two sessions
avoid downloading the same file twice, and the retention policy CRAN requires
of a package that writes to a user directory.

None of it is private. The cache is an ordinary directory tree, and that is a
deliberate property: copying it to an offline machine, restoring it from a CI
cache action, or looking at it with a file browser all work.

## Where it lives

```{r}
getaca_cache_dir()
```

That is the sandbox this vignette runs in. The default is
`tools::R_user_dir("getaca", "cache")`, the location CRAN permits for cached
downloads. Two overrides take precedence, in this order:

```{r, eval = FALSE}
options(getaca.cache = "/mnt/fast/getaca")   # this session
Sys.setenv(GETACA_CACHE = "/mnt/fast/getaca") # this process and its children
```

The environment variable is the one to reach for in CI and in job scripts,
because it survives into the R processes a build step spawns. Asking for the
directory does not create it; the first successful retrieval does.

## Layout

```
<cache>/
  blobs/sha256/<aa>/<sha256>     verified bytes, named by their own checksum
  .locks/                        one per checksum, held during a transfer
  .tmp/                          in-flight downloads, never visible as cache
  <package>/
    index.rds                    provenance for this package only
    <name>/<version>/
      raw/<file>                 this slot's name for a blob
      proc-<processor-id>/       processed result, own provenance
```

Everything a package declares is scoped by declaring package, then resource
name, then version. That falls out of identity being the triple
`package / name / version`, and it buys two things. Two packages declaring a
resource called `"backbone"` never share a slot, so one package's registry
update cannot affect another's cached data. And a version can never be overwritten by
another version, so holding two releases side by side is the normal state
rather than a special case.

The bytes underneath are shared. A file lives once, at `blobs/sha256/`, under
its own checksum, and the version slot holds a name for it: a hardlink where
the filesystem allows one, a symlink or a copy where it does not. Two packages
declaring the same 4 GB file therefore keep one copy and two independent
dependency records. They also transfer it once, because the lock is keyed on
the checksum, so the second session waits for the first and then finds the
bytes already there.

A resource composed from `part()` records stores its pieces in the same place. Each
part is admitted under its own digest, so a base every version of a series
declares is one blob however many versions declare it, and publishing a version
costs a consumer the delta rather than the whole artefact. No version slot names
a part. What reaches it is the entry composed from it, which records the part
digests beside its provenance, and that is what keeps a base alive for exactly
as long as the last cached version still holding it.

Everything the cache owns is read-only. A caller writing to a returned path
would otherwise damage every package that shares those bytes, so the write
fails at the point of the mistake instead. A caller that needs a writable
layout declares a `processor()`, which gets its own slot.

The store keeps no metadata of its own. Whether a blob is still needed is
answered by reading the package indexes, so there is no reference count that a
crash, a restored backup or a hand-deleted directory could leave disagreeing
with them.

The processed result of a processor sits beside the raw artefact rather than
replacing it, under a directory named for the processor id. Changing the
transformation means changing the id, which invalidates the derived tree
without touching the download it came from.

`index.rds` is one small file per package holding the provenance records. It
is written to a sibling temporary file and renamed, so a reader never observes
a half-written index. A metadata database was considered and left out: the
volume is tiny, and an atomic per-package file removes a dependency and a
class of locking problems.

## What a cached copy has to prove

Three different questions, three answers, and the entry record keeps them
apart so that "verified" never quietly means "we looked at this sometime".

| | what it does | when it runs | recorded as |
|---|---|---|---|
| full verification | re-hashes the bytes | on download, on `verify = TRUE`, and once the last one is older than `getaca.verify_days` | `verified_at` |
| cheap check | compares size against the entry | on every ordinary access | `checked_at` |
| access | none | on every ordinary access | `accessed_at` |

The cheap check catches truncation, replacement by a different-sized file, and
most accidental edits, for the cost of a `file.info()` call on a four-gigabyte
file. A same-size substitution passes it, which is the case the scheduled
re-hash exists to catch.

```{r, eval = FALSE}
getaca("backbone", package = "yourpkg")                 # cheap check
getaca("backbone", package = "yourpkg", verify = TRUE)  # full re-hash first
options(getaca.verify_days = 30)                  # re-hash more often
```

A cached copy that fails either check raises `getaca_error_cache_corrupt`
rather than being silently refetched, and the message names the clean-up call.
Silent repair would hide a disk going bad, and hide a colleague who edited a
file in the cache directory.

The failure is a verdict on bytes rather than on the slot that found it. Bytes
live once and every package declaring them holds its own record, so a mismatch
in the shared copy withdraws `verified_at` from every other slot naming it and
each re-hashes on next access. A slot holding its own copy, which is what a
filesystem refusing links leaves, and a processed tree derived from the bytes
are answerable only for themselves.

A part is re-hashed every time a composition reuses it, since nothing else ever
looks at it. Scheduled re-verification is driven from the entries, and a part
blob is named by an entry's part list rather than being an entry of its own, so
a base that rotted on disk would otherwise surface as the declaration failing to
produce its own artefact. A part failing its own checksum is dropped and fetched
again the way a stale partial transfer is: the declaration lists mirrors for
those bytes, and nothing else names them.

The fourth timestamp is `fetched_at`, which never moves. Together the four
answer questions that collapsing them would destroy: a resource fetched in
January, re-hashed in April and read this morning reports exactly that.

## Provenance

```{r, eval = FALSE}
getaca_info("backbone", package = "yourpkg")
#> <getaca cache entry> yourpkg/backbone@2026-06
#>   path        ~/.cache/R/getaca/yourpkg/backbone/2026-06/raw/backbone-2026-06.zip
#>   sha256      9f9f9f...
#>   size        797,000,000 bytes
#>   license     CC-BY-4.0
#>   built from  source_release: 2026-06
#>   resolved by current registry sha256:8b31e0da54cf (published 2026-07-22)
#>   source url  https://host.invalid/backbone-2026-06.zip
#>   getaca      0.0.0.9000
#>   fetched     2026-07-26 11:02:13
#>   verified    2026-07-26 11:09:44 (full re-hash)
#>   checked     2026-07-26 15:31:02 (size and mtime)
```

A composed resource reports what produced it, under the source url its series
leaves empty:

```{r, eval = FALSE}
#>   source url  NA
#>   composed    3 parts via 'concat'
#>     part      919191919191
#>     part      4e4e4e4e4e4e
#>     part      777777777777
```

An uncached resource gives `NULL`, which is what makes the call safe in a
report covering a machine that holds some of the set:

```{r}
is.null(getaca_info("backbone", registry = reg))
```

`getaca_catalogue()` is the same information across everything, plus the
declarations that have never been downloaded:

```{r}
getaca_catalogue(registry = reg)[, c("package", "name", "version",
                                     "current", "declared", "cached")]
```

With no arguments it covers every installed package that ships a registry
together with every package holding cached resources, which is the report
worth pasting into an issue:

```{r}
str(getaca_catalogue(), max.level = 1)
```

The columns worth knowing: `size` in bytes, `license`, `source` and
`registry_digest` naming the policy and the registry state that resolved it,
`parts` for how many pieces the artefact was composed from and `0` where it was
served whole, the three timestamps, `pinned`, and `path`.

## Two sessions, one download

Two R sessions asking for the same four-gigabyte file must not both fetch it,
and must never mistake each other's in-flight temporary file for a finished
resource.

The lock is a directory under `.locks/`, named for the declared checksum.
`dir.create()` is atomic on both POSIX and Windows, which makes a directory a
portable mutex with no compiled dependency and no lockfile library.

Keying it on the checksum rather than on the resource triple means the two
sessions need not be asking on behalf of the same package. Two packages
declaring the same file are waiting for the same transfer, and the one that
waits finds the bytes in the store when it wakes.

What a second session does:

1. tries to create the lock directory, and fails

2. checks whether the lock is stale, by the age of the holder file inside it

3. waits, polling, until the holder releases

4. re-reads the cache index, finds the first session's entry, and returns that
   path

Step 4 is the point. The waiter reads the entry the first session wrote and
returns that path, so the second transfer never starts. The cache check is
repeated after the lock is acquired precisely because the situation may have
changed while waiting.

Composing from parts takes one lock per distinct part digest, held for the whole
composition rather than for each transfer. Two sessions must not both fetch one
part, which is what the acquisition lock already does for a whole file. And a
part blob is named by no index until the entry composed from it is written, so
the lock is also what tells another session that bytes nothing references yet
are wanted: the `unreferenced` sweep treats a blob under an active lock as live.

A lock whose holder died leaves a directory behind. It goes stale after
`getaca.lock_stale_seconds`, defaulting to 1800, after which the next session
removes it and takes over. A session that waits longer than its timeout gets
an error naming the lock path and the `unlink()` call that clears it, so a
genuinely wedged lock is a one-line fix rather than a support thread.

```{r, eval = FALSE}
options(getaca.lock_stale_seconds = 600)
```

Set it lower for short downloads on a shared machine, higher when a single
transfer legitimately runs for an hour.

## How bytes get in

Transfers land in `.tmp/`, are sized, hashed, and only then moved into place.
An interrupted transfer can never appear as a valid cached resource, and a
failed transfer never touches a copy that was already good.

The temporary file is named after the declared checksum and the mirror that
produced it. Naming it after the checksum makes an interrupted download
resumable on the next attempt, which matters when the resource is measured in
gigabytes. Giving each mirror its own file matters for a subtler reason: a
partial transfer is resumable only against the host that produced it, so
sharing one file across mirrors would let a failed attempt at the first be
resumed onto by the second, and the resulting corruption is indistinguishable
from the publisher having changed the bytes.

A resumed transfer that completes but does not verify indicts the partial file
rather than the publisher, so the same mirror is asked once more from empty
before any conclusion is drawn about upstream. Without that, one stale
temporary file makes a resource permanently unfetchable and blames the wrong
party for it.

Verified bytes are then admitted to the store under their own checksum, and the
version slot is given a name for them. Admission is a rename, since `.tmp/` and
`blobs/` share the cache root and therefore share a filesystem. Bytes already
in the store are already named by their checksum, so admitting the same file a
second time is a no-op and the temporary copy is dropped.

A composed resource joins at that same point. Each part lands in `.tmp/` and is
admitted under its own digest, the series is combined into a second temporary
file named after the artefact's checksum, and that file is hashed against the
record before anything else sees it. Only a result matching the declaration is
admitted and given its view, so a `combiner()` needs no more trust than a mirror
does: it cannot produce bytes the declaration did not already name. A part
already in the store is used where it lies, which is the whole point of
declaring one, and an interrupted composition is overwritten by the retry rather
than accumulating.

## Watching a transfer

getaca drives its own transfer loop, so what a download looks like is a setting
rather than whatever the transfer library prints.

```{r, eval = FALSE}
getaca_progress("bar")    # redraws one line, the default when interactive
getaca_progress("line")   # one line to start and one to finish, for a log
getaca_progress("none")   # nothing
```

```
yourpkg/backbone@2026-09  [============>      ]  63%  512 MB / 812 MB  41 MB/s  ETA 00:07
```

The share is measured against the size the registry declares, which is known
before the first byte arrives and stays right when a mirror sends no content
length. A resource composed from parts reports each piece under its own label,
so a series reads as one download in stages:

```
yourpkg/backbone@2026-09 (part 1 of 3)  [===================] 100%  797 MB in 00:19
yourpkg/backbone@2026-09 (part 2 of 3)  [========>          ]  44%  4.0 MB / 9.1 MB  ...
```

`quiet = TRUE` on a single call reports nothing whatever the session is set to,
so one silent retrieval never needs the setting changed and put back:

```{r, eval = FALSE}
getaca("backbone", package = "yourpkg", quiet = TRUE)
```

A package that wants a download to look like its own writes a `reporter()`,
which is a function of one argument:

```{r}
counter <- reporter("counter", function(event) {
  if (identical(event$type, "end") && identical(event$status, "ok")) {
    message(format(event$id), ": ", event$bytes, " bytes")
  }
})
counter
```

The events are `begin`, `bytes` and `end`, and a handler switches on
`event$type` and ignores what it does not use. `begin` carries the resource, the
mirror, the declared `total` and the `offset` an interrupted transfer resumed
from; `bytes` carries the cumulative count; `end` carries the outcome. See
`?"getaca-progress"` for the fields.

A reporter never decides whether a retrieval succeeds. One that raises is
caught, reported once as a warning, and switched off for the rest of the call.

## Retention

CRAN permits `tools::R_user_dir()` on condition that contents are "actively
managed (including removing outdated material)". `getaca` reads that as a
retention policy rather than a function users might discover, so collection
runs automatically after every successful retrieval.

Five sweeps, cheapest and safest first:

| Sweep | Removes | Clock |
|---|---|---|
| `broken` | entries whose path is missing or fails the cheap check | none |
| `temp` | abandoned transfers in `.tmp/` | 7 days by mtime |
| `superseded` | unpinned versions the registry no longer names | `getaca.supersede_days`, default 30 |
| `lru` | least recently used unpinned entries | only above `getaca.max_bytes`, default 20 GB |
| `unreferenced` | bytes in the store that no entry names any more | none |

The first four sweeps remove names. `unreferenced` runs last and removes the
bytes those names were for, once the last one is gone. A blob under an active
lock is left alone: it belongs to a session that has admitted it and has not
yet written its entry.

What an entry names, for this purpose, is the blob it holds together with the
parts that blob was composed from. Reachability runs over both, so a base stays
for as long as some cached version is still composed from it and goes with the
last one.

The size ceiling measures what the cache occupies, so shared bytes count once.
Two packages declaring the same 4 GB file count 4 GB against
`getaca.max_bytes`, and evicting one of them frees nothing until the other goes
too. Part blobs are occupancy on the same terms: evicting one version of a
series reclaims the pieces no surviving version holds, and leaves a base that
another version still declares.

Superseded and not-recently-used age on separate clocks on purpose. An
expensive resource that is still the current version is never dropped merely
for being old; it is dropped only when the cache is over its ceiling, and then
only after everything already useless has gone.

Three things are never removed: pinned entries, the version the registry
currently names, and anything under an active lock.

The automatic pass after a retrieval runs only `broken` and `temp`, the two
sweeps that can only ever remove material which is already useless. Reclaiming
a superseded four-gigabyte version is a decision, so it happens on a schedule
rather than as a side effect of a download.

## Cleaning by hand

```{r}
getaca_clean(dry_run = TRUE)
```

An empty result on a fresh cache. On a working one, each row names the
package, the resource, the reason and the bytes it would reclaim, which is the
report to read before running it for real.

```{r, eval = FALSE}
getaca_clean(dry_run = TRUE)
#>   package                 resource                     reason    bytes
#> 1 yourpkg yourpkg/backbone@2026-03 superseded version past... 7.97e+08
#> 2    <NA>                     <NA>         abandoned transfer 1.20e+07

getaca_clean()                                 # run every sweep
getaca_clean(what = "temp")                    # just the abandoned transfers
getaca_clean(package = "yourpkg")              # one package
getaca_clean("backbone", package = "yourpkg")  # one resource name
```

Keeping something the sweeps would otherwise take:

```{r, eval = FALSE}
getaca_keep("backbone", package = "yourpkg")                  # pin it
getaca_keep("backbone", package = "yourpkg", pinned = FALSE)  # release the pin
```

A pinned entry is exempt from the superseded and LRU sweeps permanently. It is
still subject to the `broken` sweep, because an entry whose bytes are gone is
not worth protecting.

## Settings

Every setting is readable from an option or an environment variable, with the
option taking precedence.

| Option | Environment variable | Default | Controls |
|---|---|---|---|
| `getaca.cache` | `GETACA_CACHE` | `R_user_dir()` | where everything lives |
| `getaca.policy` | `GETACA_POLICY` | registry default | which channel resolves |
| `getaca.progress` | `GETACA_PROGRESS` | `auto` | what a transfer looks like |
| `getaca.verify_days` | `GETACA_VERIFY_DAYS` | 90 | scheduled re-hash interval |
| `getaca.supersede_days` | `GETACA_SUPERSEDE_DAYS` | 30 | retention for undeclared versions |
| `getaca.max_bytes` | `GETACA_MAX_BYTES` | 20 GB | ceiling above which LRU runs |
| `getaca.timeout` | `GETACA_TIMEOUT` | 3600 | transfer timeout in seconds |
| `getaca.lock_stale_seconds` | `GETACA_LOCK_STALE_SECONDS` | 1800 | when a lock is abandoned |
| `getaca.pin_file` | | `getaca.pins.rds` in the working directory | where pins are read from |

The defaults suit a laptop holding a couple of large reference datasets. Two
are worth revisiting on a shared machine: raise `getaca.max_bytes` when the
cache lives on a volume sized for it, and lower `getaca.lock_stale_seconds`
when transfers are short and a wedged lock costs more than a rare duplicate
download.

```{r, eval = FALSE}
options(
  getaca.cache        = "/mnt/data/getaca",
  getaca.max_bytes    = 200 * 1024^3,
  getaca.verify_days  = 30
)
```

## Moving a cache

Copy the directory. There are no absolute paths recorded inside it, so a cache
built on one machine works on another:

```{r, eval = FALSE}
# on a connected machine
Sys.setenv(GETACA_CACHE = "/tmp/seed")
getaca_prefetch(package = "yourpkg")

# then, on the machine that has no network
Sys.setenv(GETACA_CACHE = "/opt/getaca")
getaca_catalogue()   # everything already there
```

The same property is what makes CI caching work: the archive an actions cache
restores is the cache, with nothing to rebuild.

## Where to go next

- `vignette("checks")` for seeding a cache in CI

- `vignette("failures")` for `getaca_error_cache_corrupt` and its neighbours

- `vignette("policies")` for what decides which version lands in the cache

- `vignette("declaring")` for declaring a resource that arrives as a series
