---
title: "Declaring Resources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Declaring Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(getaca)

options(getaca.cache = file.path(tempdir(), "getaca-declaring"))
Sys.setenv(GETACA_OFFLINE = "true", NOT_CRAN = "true")
```

For package authors: what to declare, where to put it, and which policy to
choose. The reader here is someone whose package needs a file that is too
large, too fast-moving, or too awkwardly licensed to sit in `data/`.

Everything a declaring package owes `getaca` is one serialised object at
`inst/getaca/registry.rds`. There is no `Depends`, no `.onLoad()` hook, no
registration call, and no runtime coupling beyond calling `getaca()` when a
resource is actually needed.

## Records and channels

Two things that must not be conflated.

A **resource record** is immutable. `taxify / wfo / 2026-06` names exact bytes,
permanently. If the publisher reissues that file with different contents, that
is an upstream mutation and `getaca` refuses it.

A **channel** maps a logical name onto a record, and channels move. Where the
channel is read from is the resolution policy.

| Policy | Reads the channel from | Use when |
|---|---|---|
| `bundled` | the registry inside the installed package | default |
| `current` | the author's remote registry, falling back to bundled | data releases outpace your CRAN releases |
| `pinned` | a frozen local snapshot | an analysis must stay put |
| `offline` | nothing; cache and bundled information only | no network permitted |

`bundled` is the default because a dependency that resolves differently on
different days is not a dependency. See
`dev_notes/adr-001-registry-resolution.md` in the source for the reasoning, and
`vignette("policies")` for the operational detail.

## Anatomy of a record

```{r}
rec <- resource(
  name        = "wfo",
  version     = "2026-06",
  urls        = c("https://zenodo.invalid/records/1234567/files/wfo-2026-06.zip",
                  "https://mirror.invalid/wfo-2026-06.zip"),
  sha256      = strrep("9f", 32),
  size        = 797e6,
  license     = "CC-BY-4.0",
  description = "World Flora Online taxonomic backbone, June 2026 release"
)
rec
```

**`name`** becomes a directory name, so it is restricted to
`[A-Za-z0-9._-]`. It is scoped by your package, so a short name is fine and
`"wfo"` will never collide with another package's `"wfo"`.

**`version`** is a label under the same character restriction. `getaca` never
orders version strings, which is what lets you use whatever your upstream
actually publishes: `2026-06`, `v14.2`, `2026-06_build-3`. Pick a scheme and
keep it; the only rule the package enforces is that a label, once published,
keeps meaning the same bytes.

**`urls`** are tried in order. All must be `https://`. More than one is worth
the trouble: an outage at your primary becomes a slower first call rather than
a support issue.

**`sha256`** is the point of the whole design. It is the digest of the file as
served, which is not always the digest of the file you built: if your host
recompresses, or you upload a re-zipped copy, compute the hash after the
upload by downloading it back.

**`size`** in bytes is optional and cheap. A transfer that ends short is
detected before a multi-gigabyte hash runs, and reported as a truncation
rather than as a checksum mismatch, which points the user at a retry instead
of at the publisher.

**`license`** travels into every provenance record, so a downstream result can
report what terms its inputs came under.

**`description`** is one line for humans reading `getaca_catalogue()` output.

Malformed records are refused where they are written, with the problem named:

```{r, error = TRUE}
resource("wfo", "2026-06", urls = "http://insecure.invalid/wfo.zip",
         sha256 = strrep("9f", 32))
```

```{r, error = TRUE}
resource("wfo", "2026-06", urls = "https://ok.invalid/wfo.zip",
         sha256 = "not-a-digest")
```

## Assembling the registry

```{r}
reg <- registry(
  package   = "taxify",
  resources = list(rec)
)
reg
```

The digest on the second line identifies this registry state. It is derived
from the declaration itself, so there is no number to keep in step: change a
checksum, add a mirror or move a channel head and the digest follows. It is
recorded in the provenance of everything resolved through this registry, so a
bug report can name the exact declaration that chose the bytes.

```{r}
registry_digest(reg)
```

`registry_manifest()` shows the text that digest is taken over, which is what
makes two registries diffable when they disagree:

```{r}
registry_manifest(reg)
```

Write it into the source tree from a `data-raw/` script, so the registry is
generated rather than hand-maintained:

```{r, eval = FALSE}
# data-raw/registry.R
registry_write(reg, "inst/getaca/registry.rds")
```

Add `^data-raw$` to `.Rbuildignore` and the script stays out of the built
package while the object it produces ships.

Reading it back is symmetric, and validates:

```{r}
path <- file.path(tempdir(), "registry.rds")
registry_write(reg, path)
identical(registry_digest(registry_read(path)), registry_digest(reg))
```

Writing stamps `created`, because publishing a state is what dates it. That
date is what orders two states in time, which a digest cannot do: a digest
says whether two registries are the same, and `created` says which came
first. It sits outside the digest, so writing an unchanged registry out again
leaves its identity alone. Pass a fixed `created` when a build has to be
byte-reproducible.

```{r}
registry_read(path)$created
```

## Naming the channel head

A registry that offers one version per name has nothing to decide. One that
offers several has to say which one a bare request returns:

```{r}
reg2 <- registry(
  package = "taxify",
  current = c(wfo = "2026-09"),
  resources = list(
    resource("wfo", "2026-06", urls = "https://host.invalid/wfo-2026-06.zip",
             sha256 = strrep("9f", 32), license = "CC-BY-4.0"),
    resource("wfo", "2026-09", urls = "https://host.invalid/wfo-2026-09.zip",
             sha256 = strrep("ab", 32), license = "CC-BY-4.0")
  )
)
reg2
```

The head is marked when the registry prints, and appears as a column in the
catalogue:

```{r}
getaca_catalogue(registry = reg2)[, c("name", "version", "current", "declared")]
```

Leaving it out is an error rather than a default, because the failure it
prevents is silent. A registry appending `2026-03` after `2026-09` under a
declaration-order rule moves every user backwards without any of them
noticing:

```{r, error = TRUE}
registry(
  package = "taxify",
  resources = list(
    resource("wfo", "2026-09", urls = "https://host.invalid/a",
             sha256 = strrep("ab", 32)),
    resource("wfo", "2026-03", urls = "https://host.invalid/b",
             sha256 = strrep("cd", 32))
  )
)
```

The head names a version, so moving it is one edit and the old record stays
resolvable by `getaca(..., version = "2026-06")` for as long as you keep
declaring it. Dropping a record from the registry does not delete anyone's
cached copy; it marks it undeclared, which is what makes it eligible for the
superseded sweep after the retention window.

## Choosing `current`

Set it once, in the registry, if your data genuinely release on their own
schedule:

```{r}
remote_reg <- registry(
  package = "taxify",
  policy  = "current",
  remote  = "https://gcol33.github.io/taxify/getaca-registry.rds",
  resources = list(
    resource("wfo", "2026-06",
             urls   = "https://primary.invalid/wfo-2026-06.zip",
             sha256 = strrep("9f", 32),
             license = "CC-BY-4.0")
  )
)
```

The remote file is a static artefact on GitHub Pages, r-universe or
institutional hosting. An unreachable one falls back to the bundled registry
with a message, which is what CRAN requires when an Internet resource is
unavailable, so the hosting bar is low enough that a `docs/` directory clears
it.

What the remote channel may do:

- repair or add mirrors for a record that already exists

- publish new versions, and move the channel head onto one

What it may not do: change the checksum attached to a version that already
exists. That is rejected as an invalid registry, attributed to you rather than
to the publisher, because a remote file that can silently redefine published
bytes would undo the guarantee the whole design exists to provide.

Publishing a remote registry is the same call with a different destination:

```{r, eval = FALSE}
# In a release script, after adding the new record
registry_write(reg2, "docs/getaca-registry.rds")   # served by GitHub Pages
```

Users on `bundled` are unaffected until they reinstall; users on `current`
pick it up on their next session, or immediately after `getaca_refresh()`.

## Derived artefacts

Publishing a prepared database rather than the raw upstream release is often
the better deal for users, who then skip the expensive preprocessing. Record
both identities so provenance keeps them:

```{r}
resource(
  name    = "wfo-db",
  version = "source-2026-06_build-3",
  urls    = "https://example.invalid/wfo-db-3.duckdb",
  sha256  = strrep("ab", 32),
  license = "CC-BY-4.0",
  upstream = list(
    wfo_release    = "2026-06",
    taxifydb_build = "3"
  )
)
```

`upstream` is a free-form named list. Every entry is printed by
`getaca_info()` and kept in the cache index, so a user asking "which WFO
release is this, and which build turned it into a database" gets both answers
from one call. Two things move independently here: upstream can publish
`2026-09`, and you can fix a build bug against `2026-06`. Version labels like
`source-2026-06_build-3` keep both visible in the identity itself.

`getaca` does not care whether bytes are an official release or something you
built. It retrieves and verifies what you declared; your package owns their
scientific meaning.

## Processors

A processor turns one verified path into another. It carries a stable id, so
the derived result gets its own cache slot and its own provenance rather than
being confused with the archive it came from.

```{r}
unpack <- processor("unzip", function(input, output_dir) {
  utils::unzip(input, exdir = output_dir)
  output_dir
})
unpack$id
```

Attach it to the record that needs it:

```{r}
archive <- resource(
  name      = "wfo",
  version   = "2026-06",
  urls      = "https://host.invalid/wfo-2026-06.zip",
  sha256    = strrep("9f", 32),
  license   = "CC-BY-4.0",
  processor = unpack
)
archive
```

The function receives the verified input path and a staging directory, and
returns a path inside that directory. `getaca` renames the staging directory
into place once the function returns, so a processor that fails part-way
leaves no half-built result. A returned path outside the output directory is
an error, since the cache would then hold an entry pointing somewhere it does
not own.

Changing what the transformation does means changing the id. `"unzip"` to
`"unzip-v2"` invalidates previously processed copies without touching the raw
artefact they were built from, so users re-run the transformation rather than
re-downloading gigabytes. `getaca(..., processed = FALSE)` returns the raw
artefact for anyone who wants it.

Keep processors cheap and pure. They run inside the per-resource lock, so a
processor that takes ten minutes blocks a second session for ten minutes. If
the transformation is expensive enough to matter, publishing the derived
artefact as its own record is usually the better trade.

## Authoring in YAML

The stored form is an R object, so no parser is ever a hard dependency. YAML
and JSON are accepted at authoring time and gated at call time:

```{r, eval = FALSE}
reg <- as_registry("data-raw/resources.yml", package = "taxify")
registry_write(reg, "inst/getaca/registry.rds")
```

```yaml
wfo:
  version: "2026-06"
  urls:
    - https://primary.invalid/wfo-2026-06.zip

    - https://mirror.invalid/wfo-2026-06.zip
  sha256: "9f9f..."
  size: 797000000
  license: CC-BY-4.0
  description: World Flora Online backbone, June 2026

col:
  version: "2026-06"
  url: https://primary.invalid/col-2026-06.zip
  sha256: "ab12..."
  license: CC-BY-4.0
```

Either `urls` or `url` is accepted, and either `license` or `licence`, since
a YAML file is input rather than API. The model has one spelling for each.

The authoring format keys on resource name, so it describes one version per
name. A registry that offers several versions of a name is built in R, where
the list is a list and `current` sits beside it. Arguments passed to
`as_registry()` reach `registry()`:

```{r, eval = FALSE}
as_registry("data-raw/resources.yml", package = "taxify",
            policy = "current",
            remote = "https://gcol33.github.io/taxify/getaca-registry.rds")
```

JSON works the same way through `jsonlite`. Both are `Suggests`, and asking
for one you have not installed produces an install instruction rather than a
missing-object error.

## Handing errors to your users

Every failure is classed, and carries an `actor` field naming who can act.
Catch the ones your users will meet and say something domain-specific:

```{r, eval = FALSE}
install_backbone <- function(name = "wfo") {
  path <- tryCatch(
    getaca(name, package = "taxify"),
    getaca_error_unavailable = function(e) {
      stop("The WFO backbone is not installed and cannot be downloaded ",
           "because no network connection is available.\n",
           "Connect to the internet and run:\n",
           "  taxify::install_backbone(\"wfo\")", call. = FALSE)
    }
  )
  open_backbone(path)
}
```

`getaca_error_offline` is a subclass of `getaca_error_unavailable`, so the
handler above catches both, and a narrower handler can separate them.

Two conditions name you rather than your user.
`getaca_error_invalid_registry` means the declaration is malformed or
internally inconsistent, and `getaca_error_declaration` means several
independent mirrors agreed with each other and disagreed with your checksum.
Neither is worth catching in your own package: they are bug reports, and the
default message already says so. `vignette("failures")` covers the full set.

## Hosting

Immutable hosting makes all of this easier, and Zenodo gives it away: a
versioned DOI resolves to bytes that cannot change under you. Deposit the
data, pin the version DOI's file URL, and the upstream-mutation case stops
being possible for the copy you control.

GitHub Releases work as a second mirror. Release assets are stable once
uploaded, they are served from a CDN, and the tag names a version in a way
that reads well beside the record's own label.

`getaca` has no Zenodo client and will not grow one. It is a recommendation
about where to put files, not a feature.

## Testing your declaration

The registry is an ordinary object, so the parts worth testing are testable
without a network:

```{r, eval = FALSE}
test_that("the shipped registry is valid and names a head", {
  reg <- registry_for("taxify")
  expect_s3_class(reg, "getaca_registry")
  expect_equal(resolve_resource("wfo", registry = reg)$id$version, "2026-06")
})

test_that("the backbone parses when it is available", {
  getaca_skip_if_unavailable("wfo", package = "taxify")
  expect_s3_class(read_backbone(getaca("wfo", package = "taxify")), "backbone")
})
```

The first runs everywhere, including on CRAN, because it never leaves the
installed package. The second skips cleanly when the resource is absent and
names how to prefetch it.

Registries are read once per session and cached, so a test that installs a
registry into a temporary library needs `getaca_refresh()` to make the next
lookup take effect.

## Before you ship

- [ ] `inst/getaca/registry.rds` written by `registry_write()` from a
      `data-raw/` script
- [ ] every checksum computed from the file as served, not from a local copy
      you built
- [ ] `size` recorded, so truncation is caught before a full re-hash

- [ ] at least one mirror, ideally on a host you control

- [ ] `current` set for any name that declares more than one version

- [ ] `registry_digest()` changed since the last release, if the declaration did

- [ ] `license` on every record

- [ ] tests use `getaca_skip_if_unavailable()`

- [ ] examples use `getaca_optional()`, or `\donttest{}` for large resources

- [ ] `R CMD check` passes with no network available
