---
title: "Declaring Resources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Declaring Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(getaca)

options(getaca.cache = file.path(tempdir(), "getaca-declaring"))
Sys.setenv(GETACA_OFFLINE = "true", NOT_CRAN = "true")
```

For package authors: what to declare, where to put it, and which policy to
choose. The reader here is someone whose package needs a file that is too
large, too fast-moving, or too awkwardly licensed to sit in `data/`.

Everything a declaring package owes `getaca` is one serialised object at
`inst/getaca/registry.rds`. There is no `Depends`, no `.onLoad()` hook, no
registration call, and no runtime coupling beyond calling `getaca()` when a
resource is actually needed.

## Records and channels

Two things that must not be conflated.

A **resource record** is immutable. `yourpkg / backbone / 2026-06` names exact
bytes, permanently. If the publisher reissues that file with different contents,
that is an upstream mutation and `getaca` refuses it.

A **channel** maps a logical name onto a record, and channels move. Where the
channel is read from is the resolution policy.

| Policy | Reads the channel from | Use when |
|---|---|---|
| `bundled` | the registry inside the installed package | default |
| `current` | the author's remote registry, falling back to bundled | data releases outpace your CRAN releases |
| `pinned` | a frozen local snapshot | an analysis must stay put |
| `offline` | nothing; cache and bundled information only | no network permitted |

`bundled` is the default because a dependency that resolves differently on
different days is not a dependency. See
`dev_notes/adr-001-registry-resolution.md` in the source for the reasoning, and
`vignette("policies")` for the operational detail.

## Anatomy of a record

```{r}
rec <- resource(
  name        = "backbone",
  version     = "2026-06",
  urls        = c("https://zenodo.invalid/records/1234567/files/backbone-2026-06.zip",
                  "https://mirror.invalid/backbone-2026-06.zip"),
  sha256      = strrep("9f", 32),
  size        = 797e6,
  license     = "CC-BY-4.0",
  description = "Reference backbone, June 2026 release"
)
rec
```

**`name`** becomes a directory name, so it is restricted to
`[A-Za-z0-9._-]`. It is scoped by your package, so a short name is fine and
`"backbone"` will never collide with another package's `"backbone"`.

**`version`** is a label under the same character restriction. `getaca` never
orders version strings, which is what lets you use whatever your upstream
actually publishes: `2026-06`, `v14.2`, `2026-06_build-3`. Pick a scheme and
keep it; the only rule the package enforces is that a label, once published,
keeps meaning the same bytes.

**`urls`** are tried in order. All must be `https://`. More than one is worth
the trouble: an outage at your primary becomes a slower first call rather than
a support issue.

**`sha256`** is the point of the whole design. It is the digest of the file as
served, which is not always the digest of the file you built: if your host
recompresses, or you upload a re-zipped copy, compute the hash after the
upload by downloading it back. `registry_draft()`, below, does exactly that.

**`size`** in bytes is optional and cheap. A transfer that ends short is
detected before a multi-gigabyte hash runs, and reported as a truncation
rather than as a checksum mismatch, which points the user at a retry instead
of at the publisher.

**`license`** travels into every provenance record, so a downstream result can
report what terms its inputs came under.

**`description`** is one line for humans reading `getaca_catalogue()` output.

**`file`** is optional and names the cached file. The default is the file name
in the first URL, which is right until the URL ends in `download?id=7`, and
which a record composed from parts has to state for itself.

Malformed records are refused where they are written, with the problem named:

```{r, error = TRUE}
resource("backbone", "2026-06",
         urls = "http://insecure.invalid/backbone.zip",
         sha256 = strrep("9f", 32))
```

```{r, error = TRUE}
resource("backbone", "2026-06",
         urls = "https://ok.invalid/backbone.zip",
         sha256 = "not-a-digest")
```

## Drafting from where the data is

Typing a record by hand is fine for everything except the checksum, which has
to come from the bytes. `registry_draft()` takes the locations, retrieves each
file once, hashes it locally and returns a registry:

```{r, eval = FALSE}
reg <- registry_draft(
  c(backbone = "https://zenodo.org/records/1234567/files/backbone-2026-06.zip"),
  package = "yourpkg",
  version = "2026-06"
)
```

A location is a plain URL, or an identifier for an archive holding several
files. Which one it is, and which archive, is read off the string:

```{r, eval = FALSE}
registry_draft("10.5281/zenodo.4924875", package = "yourpkg")
registry_draft("10.6084/m9.figshare.14763051.v1", package = "yourpkg")
registry_draft("10.11588/data/TKCFEF", package = "yourpkg")
```

Zenodo and figshare register their own DOI prefixes, so those are recognised
from the string with no lookup. A Dataverse instance is self-hosted under its
own prefix, so a bare DOI is resolved through `doi.org` to find which host to
ask. Each archive supplies what it knows: the licence, the version, and a DOI
for the artefact. Zenodo mints a record id per version and that becomes the
version label; figshare and Dataverse carry their own version numbers. For a
plain URL there is nothing to infer, so `version =` is required.

Checksums the archive reports are read past. All three publish md5, and all
three publish it from the host that serves the file, so it says nothing the
transfer has not already said. What ends up in the record is a SHA-256 computed
here, from the bytes that arrived.

A draft is a starting point. Resources are named after their files, which is
rarely the name you want a user to type, and `description` is empty. Both are
ordinary `resource()` arguments: edit the returned registry, or write the
`registry_draft()` call into `data-raw/` and edit around it. Give a name in the
call where a location is one file:

```{r, eval = FALSE}
reg <- registry_draft(
  list(backbone = c("https://zenodo.org/records/1234567/files/backbone-2026-06.zip",
               "https://mirror.invalid/backbone-2026-06.zip")),
  package = "yourpkg",
  version = "2026-06"
)
registry_write(reg, "inst/getaca/registry.rds")
```

A list element is the mirrors of one resource. The first is retrieved and
hashed; the rest are recorded and walked at fetch time like any other mirror
list.

A checksum can only come from the bytes, but drafting a large record costs no
disk: the file is hashed as it arrives and never written down. Pass
`keep = TRUE` to write it to the cache instead, so the first `getaca()` call
for the drafted resource finds it already there.

Where you have the file already, `local =` hashes the copy on your machine and
transfers nothing. The record still names the location, since that is where
users fetch from:

```{r, eval = FALSE}
registry_draft(
  c(backbone = "https://zenodo.org/records/1234567/files/backbone-2026-06.zip"),
  package = "yourpkg", version = "2026-06",
  local   = c(backbone = "~/build/backbone-2026-06.zip")
)
```

That records your build rather than the download, which are the same bytes
only if the host served back what it was given. `sha256 =` alongside it holds
one to the other, and the draft fails where they differ. `sha256 =` on its own
declares a checksum you already hold and retrieves nothing at all.

## Assembling the registry

```{r}
reg <- registry(
  package   = "yourpkg",
  resources = list(rec)
)
reg
```

The digest on the second line identifies this registry state. It is derived
from the declaration itself, so there is no number to keep in step: change a
checksum, add a mirror or move a channel head and the digest follows. It is
recorded in the provenance of everything resolved through this registry, so a
bug report can name the exact declaration that chose the bytes.

```{r}
registry_digest(reg)
```

`registry_manifest()` shows the text that digest is taken over, which is what
makes two registries diffable when they disagree:

```{r}
registry_manifest(reg)
```

Write it into the source tree from a `data-raw/` script, so the registry is
generated rather than hand-maintained:

```{r, eval = FALSE}
# data-raw/registry.R
registry_write(reg, "inst/getaca/registry.rds")
```

Add `^data-raw$` to `.Rbuildignore` and the script stays out of the built
package while the object it produces ships.

Reading it back is symmetric, and validates:

```{r}
path <- file.path(tempdir(), "registry.rds")
registry_write(reg, path)
identical(registry_digest(registry_read(path)), registry_digest(reg))
```

Writing stamps `created`, because publishing a state is what dates it. That
date is what orders two states in time, which a digest cannot do: a digest
says whether two registries are the same, and `created` says which came
first. It sits outside the digest, so writing an unchanged registry out again
leaves its identity alone. Pass a fixed `created` when a build has to be
byte-reproducible.

```{r}
registry_read(path)$created
```

## Naming the channel head

A registry that offers one version per name has nothing to decide. One that
offers several has to say which one a bare request returns:

```{r}
reg2 <- registry(
  package = "yourpkg",
  current = c(backbone = "2026-09"),
  resources = list(
    resource("backbone", "2026-06",
             urls = "https://host.invalid/backbone-2026-06.zip",
             sha256 = strrep("9f", 32), license = "CC-BY-4.0"),
    resource("backbone", "2026-09",
             urls = "https://host.invalid/backbone-2026-09.zip",
             sha256 = strrep("ab", 32), license = "CC-BY-4.0")
  )
)
reg2
```

The head is marked when the registry prints, and appears as a column in the
catalogue:

```{r}
getaca_catalogue(registry = reg2)[, c("name", "version", "current", "declared")]
```

Leaving it out is an error rather than a default, because the failure it
prevents is silent. A registry appending `2026-03` after `2026-09` under a
declaration-order rule moves every user backwards without any of them
noticing:

```{r, error = TRUE}
registry(
  package = "yourpkg",
  resources = list(
    resource("backbone", "2026-09",
             urls = "https://host.invalid/a",
             sha256 = strrep("ab", 32)),
    resource("backbone", "2026-03",
             urls = "https://host.invalid/b",
             sha256 = strrep("cd", 32))
  )
)
```

The head names a version, so moving it is one edit and the old record stays
resolvable by `getaca(..., version = "2026-06")` for as long as you keep
declaring it. Dropping a record from the registry does not delete anyone's
cached copy; it marks it undeclared, which is what makes it eligible for the
superseded sweep after the retention window.

## Choosing `current`

Set it once, in the registry, if your data genuinely release on their own
schedule:

```{r}
remote_reg <- registry(
  package = "yourpkg",
  policy  = "current",
  remote  = "https://yourpkg.invalid/getaca-registry.rds",
  resources = list(
    resource("backbone", "2026-06",
             urls   = "https://primary.invalid/backbone-2026-06.zip",
             sha256 = strrep("9f", 32),
             license = "CC-BY-4.0")
  )
)
```

The remote file is a static artefact on GitHub Pages, r-universe or
institutional hosting. An unreachable one falls back to the bundled registry
with a message, which is what CRAN requires when an Internet resource is
unavailable, so the hosting bar is low enough that a `docs/` directory clears
it.

What the remote channel may do:

- repair or add mirrors for a record that already exists

- publish new versions, and move the channel head onto one

What it may not do: change the checksum attached to a version that already
exists. That is rejected as an invalid registry, attributed to you rather than
to the publisher, because a remote file that can silently redefine published
bytes would undo the guarantee the whole design exists to provide.

Publishing a remote registry is the same call with a different destination:

```{r, eval = FALSE}
# In a release script, after adding the new record
registry_write(reg2, "docs/getaca-registry.rds")   # served by GitHub Pages
```

Users on `bundled` are unaffected until they reinstall; users on `current`
pick it up on their next session, or immediately after `getaca_refresh()`.

## Derived artefacts

Publishing a prepared database rather than the raw upstream release is often
the better deal for users, who then skip the expensive preprocessing. Record
both identities so provenance keeps them:

```{r}
resource(
  name    = "backbone-db",
  version = "source-2026-06_build-3",
  urls    = "https://example.invalid/backbone-db-3.duckdb",
  sha256  = strrep("ab", 32),
  license = "CC-BY-4.0",
  upstream = list(
    source_release    = "2026-06",
    build = "3"
  )
)
```

`upstream` is a free-form named list. Every entry is printed by
`getaca_info()` and kept in the cache index, so a user asking "which upstream
release is this, and which build turned it into a database" gets both answers
from one call. Two things move independently here: upstream can publish
`2026-09`, and you can fix a build bug against `2026-06`. Version labels like
`source-2026-06_build-3` keep both visible in the identity itself.

`getaca` does not care whether bytes are an official release or something you
built. It retrieves and verifies what you declared; your package owns their
scientific meaning.

## Files that arrive in parts

Two situations call for a record whose bytes arrive in pieces. Your host caps
the size of a single file, so a 6 GB database goes up as chunks. Or upstream
publishes a base release and then issues deltas against it, because each delta
is a hundredth of the size of the thing it updates.

Declare the pieces with `part()` and the artefact they compose with `sha256`,
as usual:

```{r}
series <- resource(
  name    = "backbone",
  version = "2026-09",
  sha256  = strrep("b1", 32),
  size    = 812e6,
  file    = "backbone.parquet",
  license = "CC-BY-4.0",
  parts   = list(
    part("https://zenodo.invalid/records/456/files/backbone-base.bin",
         sha256 = strrep("91", 32), size = 797e6),
    part("https://zenodo.invalid/records/456/files/backbone-2026-06.bin",
         sha256 = strrep("4e", 32), size = 9.1e6),
    part("https://zenodo.invalid/records/456/files/backbone-2026-09.bin",
         sha256 = strrep("77", 32), size = 5.4e6)
  )
)
series
```

A record names locations for the whole file or the parts it is composed from,
so `urls` and `parts` are alternatives and declaring both is refused. Each part
is fetched through the same mirror walk, verified against its own checksum, and
stored under its own digest. The pieces are then combined, the result is hashed
against the record's `sha256`, and it joins the store exactly as a downloaded
file does. Everything after that point is identical: the same blob, the same
cached path, the same periodic re-verification, the same `getaca_info()`.

Storing each part under its own digest is what makes the next version cheap.
The base above is one blob, and every version declaring it points at that blob,
so `2026-12` costs your users the new delta rather than 812 MB. The trade is
disk: the parts stay for as long as some cached version still names them, which
is what keeps the saving available.

**`sha256` still describes the artefact**, which is the rule everything else
follows from. A version means the bytes its checksum names and nothing about
the route to them, so re-splitting a file at different boundaries, adding a
mirror for one piece, or moving the series to a new host are all changes of
route. Your users are held to the artefact either way, and a series that does
not produce it is reported against the declaration:

```{r, eval = FALSE}
getaca("backbone", package = "yourpkg")
#> Error: The parts declared for yourpkg/backbone@2026-09 do not produce the declared bytes.
#>   3 parts, each matching its own checksum, combined by 'concat'
#>   declared SHA-256: b1b1b1...
#>   composed SHA-256: 2c40f9...
#>
#> Every part arrived intact, so this is not a transfer problem.
```

### Combining by something other than concatenation

Parts are concatenated in declaration order, which is what a file split for an
upload limit needs. A delta format needs `combiner()`, since applying a patch
is knowledge about a file format and `getaca` has none:

```{r}
apply_deltas <- combiner("bsdiff", function(parts, output) {
  current <- parts[1]
  for (delta in parts[-1]) {
    current <- bspatch(current, delta, tempfile())
  }
  file.copy(current, output)
})
apply_deltas
```

The function receives the verified part paths in declaration order and the file
to write. Attach it with `combiner = apply_deltas`. Like a processor, it carries
an id, and the manifest records that id rather than the closure. Unlike a
processor, what it produced is checked against the record's own `sha256` before
anyone sees it, so an id is all the manifest needs: the checksum is what says
the result is right.

Order matters more here than it does for mirrors. Mirrors are tried in the order
given and any one of them ends the walk; parts are combined in the order given,
so two orderings of one series are two different artefacts, and reordering a
series changes the registry digest.

So the two situations this section opened with are the same declaration with a
different combiner. A file split for an upload limit is `part()` and the
default. A base with deltas against it is `part()` and a `combiner()` that
knows the delta format.

### A series that keeps growing

For a base with deltas, each version declares the whole series from the base
onwards: `2026-12` is the base and three deltas. Your users still transfer only
the new delta, since the earlier pieces are already blobs in their store, and
the combiner reapplies the chain from the base each time. That is local work
rather than transfer, and for most formats it is the cheaper of the two by a
wide margin.

When the chain gets long enough that reapplying it stops being cheap, publish
the composed artefact of each version alongside its deltas and anchor the next
version on it:

```{r, eval = FALSE}
resource(
  name    = "backbone",
  version = "2026-12",
  sha256  = strrep("c2", 32),
  file    = "backbone.parquet",
  parts   = list(
    # The artefact of 2026-09, at its published checksum.
    part("https://zenodo.invalid/records/456/files/backbone-2026-09.parquet",
         sha256 = strrep("b1", 32)),
    part("https://zenodo.invalid/records/789/files/backbone-2026-12.bin",
         sha256 = strrep("d3", 32))
  ),
  combiner = apply_deltas
)
```

Nothing new is needed for this, and it costs nothing extra to the users it does
not help. A machine holding `2026-09` already has those bytes under that digest,
finds them in the store, and applies one delta. A machine holding nothing
downloads the previous artefact whole and applies one delta. Which happens is
decided by what is in the store, and the record reads the same either way.

## Processors

A processor turns one verified path into another. It carries a stable id, so
the derived result gets its own cache slot and its own provenance rather than
being confused with the archive it came from.

Most declarations of an archive want the same transformation, so it is shipped.
`unpack()` reads the format from the cached file's name:

```{r}
archive <- resource(
  name      = "backbone",
  version   = "2026-06",
  urls      = "https://host.invalid/backbone-2026-06.zip",
  sha256    = strrep("9f", 32),
  license   = "CC-BY-4.0",
  processor = unpack()
)
archive
```

It covers `.zip`, the tarballs under any compression (`.tar`, `.tar.gz`,
`.tgz`, `.tar.bz2`, `.tbz2`, `.tar.xz`, `.txz`) and a single compressed file
(`.gz`, `.bz2`, `.xz`), which is decompressed under its own name with the
extension dropped. Name the format for a file whose name does not carry one:

```{r}
unpack("gzip")$id
```

A large archive holding one subtree you need takes `members`, which names a
file, or a directory and everything under it:

```{r}
unpack(members = "tables")$id
```

The id encodes the settings for a reason. It is what the cache slot and the
registry manifest are keyed on, so two records asking for different subsets of
one archive must not resolve to one slot. A name that matches nothing in the
archive is an error rather than an empty result, since getaca hashes the
archive and never looks at what came out of it: an empty slot would pass every
later check.

### Writing your own

Anything else is a function of two arguments:

```{r}
index <- processor("index-v1", function(input, output_dir) {
  out <- file.path(output_dir, "backbone.tsv")
  write.table(read.csv(input), out, sep = "\t", row.names = FALSE)
  out
})
index$id
```

The function receives the verified input path and a staging directory, and
returns a path inside that directory. `getaca` renames the staging directory
into place once the function returns, so a processor that fails part-way
leaves no half-built result. A returned path outside the output directory is
an error, since the cache would then hold an entry pointing somewhere it does
not own.

Changing what the transformation does means changing the id. `"index-v1"` to
`"index-v2"` invalidates previously processed copies without touching the raw
artefact they were built from, so users re-run the transformation rather than
re-downloading gigabytes. `getaca(..., processed = FALSE)` returns the raw
artefact for anyone who wants it.

Keep processors cheap and pure. They run inside the per-resource lock, so a
processor that takes ten minutes blocks a second session for ten minutes. If
the transformation is expensive enough to matter, publishing the derived
artefact as its own record is usually the better trade.

## Authoring in YAML

The stored form is an R object, so no parser is ever a hard dependency. YAML
and JSON are accepted at authoring time and gated at call time:

```{r, eval = FALSE}
reg <- as_registry("data-raw/resources.yml", package = "yourpkg")
registry_write(reg, "inst/getaca/registry.rds")
```

```yaml
backbone:
  version: "2026-06"
  urls:
    - https://primary.invalid/backbone-2026-06.zip

    - https://mirror.invalid/backbone-2026-06.zip
  sha256: "9f9f..."
  size: 797000000
  license: CC-BY-4.0
  description: Reference backbone, June 2026

grid:
  version: "2026-06"
  url: https://primary.invalid/grid-2026-06.zip
  sha256: "ab12..."
  license: CC-BY-4.0
```

Either `urls` or `url` is accepted, and either `license` or `licence`, since
a YAML file is input rather than API. The model has one spelling for each.

The authoring format keys on resource name, so it describes one version per
name. A registry that offers several versions of a name is built in R, where
the list is a list and `current` sits beside it. Arguments passed to
`as_registry()` reach `registry()`:

```{r, eval = FALSE}
as_registry("data-raw/resources.yml", package = "yourpkg",
            policy = "current",
            remote = "https://yourpkg.invalid/getaca-registry.rds")
```

JSON works the same way through `jsonlite`. Both are `Suggests`, and asking
for one you have not installed produces an install instruction rather than a
missing-object error.

## Handing errors to your users

Every failure is classed, and carries an `actor` field naming who can act.
Catch the ones your users will meet and say something domain-specific:

```{r, eval = FALSE}
install_backbone <- function(name = "backbone") {
  path <- tryCatch(
    getaca(name, package = "yourpkg"),
    getaca_error_unavailable = function(e) {
      stop("The backbone is not installed and cannot be downloaded ",
           "because no network connection is available.\n",
           "Connect to the internet and run:\n",
           "  yourpkg::install_backbone(\"backbone\")", call. = FALSE)
    }
  )
  open_backbone(path)
}
```

`getaca_error_offline` is a subclass of `getaca_error_unavailable`, so the
handler above catches both, and a narrower handler can separate them.

Two conditions name you rather than your user.
`getaca_error_invalid_registry` means the declaration is malformed or
internally inconsistent, and `getaca_error_declaration` means several
independent mirrors agreed with each other and disagreed with your checksum.
Neither is worth catching in your own package: they are bug reports, and the
default message already says so. `vignette("failures")` covers the full set.

## Hosting

Immutable hosting makes all of this easier, and Zenodo gives it away: a
versioned DOI resolves to bytes that cannot change under you. Deposit the
data, pin the version DOI's file URL, and the upstream-mutation case stops
being possible for the copy you control.

GitHub Releases work as a second mirror. Release assets are stable once
uploaded, they are served from a CDN, and the tag names a version in a way
that reads well beside the record's own label.

An archive is consulted when a registry is written and never when a user
fetches. `registry_draft()` is the whole of it, and what it produces is a list
of `https://` locations, which are what ship. A Zenodo file URL is derivable
from the record anyway: `10.5281/zenodo.1234567` and a file name give
`https://zenodo.org/records/1234567/files/<name>`, which is a `urls` entry like
any other. Record the identifier itself in `doi =`, where it is cited from
rather than fetched from:

```{r, eval = FALSE}
resource(
  "backbone", "2026-06",
  urls = "https://zenodo.org/records/1234567/files/backbone-2026-06.zip",
  doi = "10.5281/zenodo.1234567",
  sha256 = "...", license = "CC-BY-4.0"
)
```

The DOI travels into provenance and appears in `getaca_info()` and
`getaca_catalogue()`, so an analysis can report the identifier for the exact
bytes it read.

## Data behind a registration

Some archives serve their files only to a registered account. A declaration
says which credential a host requires; it never carries one. `bearer()` and
`basic()` name environment variables, and `auth_host()` binds a scheme to the
host it applies to:

```{r, eval = FALSE}
registry(
  package = "canopy",
  auth = list(
    auth_host("data.ornldaac.earthdata.nasa.gov",
              bearer("EARTHDATA_TOKEN"),
              register = "https://urs.earthdata.nasa.gov/users/new")
  ),
  resources = list(
    resource("gedi-l4a", "2026-06",
             urls = "https://data.ornldaac.earthdata.nasa.gov/gedi/l4a-2026-06.h5",
             sha256 = "...", license = "CC-BY-4.0")
  )
)
```

The credential is read from the environment at the moment of the request. It is
never stored, never written to the cache, never recorded in provenance and never
printed. It travels as an `Authorization` header, which libcurl withholds from a
redirect to another host, and these archives redirect to object storage
routinely. A token carried in a query string would have neither protection and
would appear in every error message listing the URLs that were tried, so one is
not accepted.

A credential belongs to a host rather than to a file, so the declaration is
matched by host: a record listing an authenticated mirror beside a public one
presents the credential to the first and not to the second, and a record that
falls through to the public mirror simply succeeds. Hosts match exactly, without
wildcards.

Credentials are read from the registry the package *ships*, never from a remote
one, for the reason signing keys are. A declaration arriving over the network
that could name the hosts a credential is sent to would be able to nominate its
own.

Say plainly in your documentation which variable a user sets and where they
register. `getaca_credentials()` reports both, and whether the variable is set,
without touching the network:

```{r, eval = FALSE}
getaca_credentials(package = "canopy")
```

The reproducibility promise narrows for these resources, from every machine
getting the same bytes to every machine holding the credential getting the same
bytes. Weigh that against a public mirror before declaring one.

## Testing your declaration

The registry is an ordinary object, so the parts worth testing are testable
without a network:

```{r, eval = FALSE}
test_that("the shipped registry is valid and names a head", {
  reg <- registry_for("yourpkg")
  expect_s3_class(reg, "getaca_registry")
  expect_equal(resolve_resource("backbone", registry = reg)$id$version, "2026-06")
})

test_that("the backbone parses when it is available", {
  getaca_skip_if_unavailable("backbone", package = "yourpkg")
  expect_s3_class(read_backbone(getaca("backbone", package = "yourpkg")), "backbone")
})
```

The first runs everywhere, including on CRAN, because it never leaves the
installed package. The second skips cleanly when the resource is absent and
names how to prefetch it.

Registries are read once per session and cached, so a test that installs a
registry into a temporary library needs `getaca_refresh()` to make the next
lookup take effect.

## Before you ship

- [ ] `inst/getaca/registry.rds` written by `registry_write()` from a
      `data-raw/` script
- [ ] every checksum computed from the file as served, not from a local copy
      you built
- [ ] `size` recorded, so truncation is caught before a full re-hash

- [ ] at least one mirror, ideally on a host you control

- [ ] `current` set for any name that declares more than one version

- [ ] `registry_digest()` changed since the last release, if the declaration did

- [ ] `license` on every record

- [ ] tests use `getaca_skip_if_unavailable()`

- [ ] examples use `getaca_optional()`, or `\donttest{}` for large resources

- [ ] `R CMD check` passes with no network available
