---
title: "Declaring Resources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Declaring Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(getaca)
```

For package authors: what to declare, where to put it, and which policy to
choose.

## Records and channels

Two things that must not be conflated.

A **resource record** is immutable. `taxify / wfo / 2026-06` names exact bytes,
permanently. If the publisher reissues that file with different contents, that
is an upstream mutation and `getaca` refuses it.

A **channel** maps a logical name onto a record, and channels move. Where the
channel is read from is the resolution policy.

| Policy | Reads the channel from | Use when |
|---|---|---|
| `bundled` | the registry inside the installed package | default |
| `current` | the author's remote registry, falling back to bundled | data releases outpace your CRAN releases |
| `pinned` | a frozen local snapshot | an analysis must stay put |
| `offline` | nothing; cache and bundled information only | no network permitted |

`bundled` is the default because a dependency that resolves differently on
different days is not a dependency. See
`dev_notes/adr-001-registry-resolution.md` in the source for the reasoning.

## Choosing `current`

Set it once, in the registry, if your data genuinely release on their own
schedule:

```{r}
reg <- registry(
  package = "taxify",
  policy  = "current",
  remote  = "https://gcol33.github.io/taxify/getaca-registry.rds",
  resources = list(
    resource("wfo", "2026-06",
             urls   = "https://primary.invalid/wfo-2026-06.zip",
             sha256 = strrep("9f", 32),
             licence = "CC-BY-4.0")
  )
)
```

The remote file is a static artefact on GitHub Pages, r-universe or
institutional hosting. It is not a service, and it does not need to be up: an
unreachable remote registry falls back to bundled with a message, because CRAN
requires graceful degradation when an Internet resource is unavailable.

What the remote channel may do:

- repair or add mirrors for a record that already exists
- publish new versions, and point the channel at one

What it may not do: change the checksum attached to a version that already
exists. That is rejected as an invalid registry, attributed to you rather than
to the publisher.

## Versions are labels

`getaca` does not order version strings, because they are not semantic
versions. `2026-06` and `source-2026-06_build-3` are labels. The channel head
is the last matching record in declaration order, so list versions oldest
first, and let users hold a specific one explicitly:

```{r, eval = FALSE}
getaca("wfo", package = "taxify", version = "2026-06")
```

## Derived artefacts

Publishing a prepared database rather than the raw upstream release is often
the better deal for users, who then skip the expensive preprocessing. Record
both identities so provenance keeps them:

```{r}
resource(
  name    = "wfo-db",
  version = "source-2026-06_build-3",
  urls    = "https://example.invalid/wfo-db-3.duckdb",
  sha256  = strrep("ab", 32),
  licence = "CC-BY-4.0",
  upstream = list(
    wfo_release    = "2026-06",
    taxifydb_build = "3"
  )
)
```

`getaca` does not care whether bytes are an official release or something you
built. It retrieves and verifies what you declared; your package owns their
scientific meaning.

## Processors

A processor turns one verified path into another. It carries a stable id, so
the derived result gets its own cache slot and its own provenance rather than
being confused with the archive it came from.

```{r}
unpack <- processor("unzip", function(input, output_dir) {
  utils::unzip(input, exdir = output_dir)
  output_dir
})
```

Changing what the transformation does means changing the id, which invalidates
previously processed copies. `getaca(..., processed = FALSE)` returns the raw
artefact.

## Authoring in YAML

The stored form is an R object, so no parser is ever a hard dependency. YAML
and JSON are accepted at authoring time and gated at call time:

```{r, eval = FALSE}
reg <- as_registry("data-raw/resources.yml", package = "taxify")
registry_write(reg, "inst/getaca/registry.rds")
```

```yaml
wfo:
  version: "2026-06"
  urls:
    - https://primary.invalid/wfo-2026-06.zip
    - https://mirror.invalid/wfo-2026-06.zip
  sha256: "9f9f..."
  licence: CC-BY-4.0
```

## Handing errors to your users

Every failure is classed, and carries an `actor` field naming who can act.
Catch the ones your users will meet and say something domain-specific:

```{r, eval = FALSE}
install_backbone <- function(name = "wfo") {
  path <- tryCatch(
    getaca(name, package = "taxify"),
    getaca_error_unavailable = function(e) {
      stop("The WFO backbone is not installed and cannot be downloaded ",
           "because no network connection is available.\n",
           "Connect to the internet and run:\n",
           "  taxify::install_backbone(\"wfo\")", call. = FALSE)
    }
  )
  open_backbone(path)
}
```

`getaca_error_offline` is a subclass of `getaca_error_unavailable`, so the
handler above catches both, and a narrower handler can separate them.

## Hosting

Immutable hosting makes all of this easier, and Zenodo gives it away: a
versioned DOI resolves to bytes that cannot change under you. Deposit the
data, pin the version DOI's file URL, and the upstream-mutation case stops
being possible.

`getaca` has no Zenodo client and will not grow one. It is a recommendation
about where to put files, not a feature.

## Before you ship

- [ ] `inst/getaca/registry.rds` written by `registry_write()`
- [ ] every checksum computed from the file as served, not from a local copy
      you built
- [ ] at least one mirror, ideally on a host you control
- [ ] tests use `getaca_skip_if_unavailable()`
- [ ] examples use `getaca_optional()`, or `\donttest{}` for large resources
- [ ] `R CMD check` passes with no network available
